### 2014-12-07 text/AppendixA.tex
Finish cbn section, notes on normal forms (Chapter 4)
Finish equational theory, subtyping and lambda (Chapter 4)
Finish tweaks in flat coeffects (Chapter 3)

%---------------------------------------------------------------------------------------------------

\section{Internalized substitution}

\subsection{First transformation}

\begin{equation*}
(\kvd{glet}~x=e_1~\kvd{in}~e_2)~e_3 \leadsto \kvd{glet}~x=e_1~\kvd{in}~(e_2~e_3)
\end{equation*}

\begin{equation*}
\tyrule{app}
  { \tyrule{glet}
      { \coctx{\Gamma}{\cclrd{s}} \vdash e_1 : \tau_1 &
        \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r}} \vdash e_2 : \tau_3 \xrightarrow{\cclrd{t}} \tau_2 }
      { \coctx{\Gamma}{\cclrd{r}\;\cpar\;({\cclrd{s}\,\cseq\,\cclrd{r})}} \vdash \kvd{glet}~x=e_1~\kvd{in}~e_2 : \tau_3 \xrightarrow{\cclrd{t}} \tau_2 } &
    \coctx{\Gamma}{\cclrd{u}} \vdash e_3 : \tau_3  }
  { \coctx{\Gamma}{(\cclrd{r}\;\cpar\;(\cclrd{s}\,\cseq\,\cclrd{r})) \;\cpar\; (\cclrd{u}\,\cseq\,\cclrd{t}) } \vdash (\kvd{glet}~x=e_1~\kvd{in}~e_2)~e_3 : \tau_2 }
\end{equation*}

to

\begin{equation*}
\tyrule{glet}
  { \coctx{\Gamma}{\cclrd{s}} \vdash e_1 : \tau_1 &
    \tyrule{app}
      { \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r}} \vdash e_2 : \tau_3 \xrightarrow{\cclrd{t}} \tau_2 & 
        \coctx{\Gamma}{\cclrd{u}} \vdash e_3 : \tau_3 }
      { \coctx{\Gamma}{\cclrd{r}\;\cpar\;(\cclrd{u}\,\cseq\,\cclrd{t})} \vdash e_2~e_3 : \tau_2 } }
  { \coctx{\Gamma}{ (\cclrd{r}\;\cpar\;(\cclrd{u}\,\cseq\,\cclrd{t})) \;\cpar\; (\cclrd{s}\,\cseq\,(\cclrd{r}\;\cpar\;(\cclrd{u}\,\cseq\,\cclrd{t})))  }
      \vdash \kvd{glet}~x=e_1~\kvd{in}~(e_2~e_3) : \tau_2 }
\end{equation*}

meaning
\begin{equation*}
\begin{array}{l}
(\cclrd{r}\;\cpar\;(\cclrd{s}\,\cseq\,\cclrd{r})) \;\cpar\; (\cclrd{u}\,\cseq\,\cclrd{t}) =
\end{array}
\end{equation*}

\subsection{Second transformation}

Second transformation

\begin{equation*}
\tyrule{glet}
  { \coctx{\Gamma}{\cclrd{s}} \vdash e_s : \tau_s &
    \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r}} \vdash e_r : \tau_r &
    \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{t}} \vdash e_t : \tau_t }
  { \coctx{\Gamma}{
      \cclrd{t}
      \;\cpar\;
      (  (\cclrd{r}\;\cpar\;(\cclrd{s}\,\cseq\,\cclrd{r}))
         \,\cseq\, 
         \cclrd{t} )
    } 
    \vdash \kvd{glet}~x_r=(\kvd{glet}~x_s=e_s~\kvd{in}~e_r)~\kvd{in}~e_t : \tau_t }
\end{equation*}

or

\begin{equation*}
\tyrule{glet}
  { \coctx{\Gamma}{\cclrd{s}} \vdash e_s : \tau_s &
    \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r}} \vdash e_r : \tau_r &
    \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{t}} \vdash e_t : \tau_t }
  { \coctx{\Gamma}{
      (\cclrd{t}\;\cpar\;(\cclrd{r}\,\cseq\,\cclrd{t})) 
      \;\cpar\;
      (  \cclrd{s}
         \,\cseq\, 
         (\cclrd{t}\;\cpar\;(\cclrd{r}\,\cseq\,\cclrd{t})) )
    } 
    \vdash \kvd{glet}~x_s=e_s~\kvd{in}~(\kvd{glet}~x_r=e_r~\kvd{in}~e_t) : \tau_t }
\end{equation*}

\begin{equation*}
\begin{array}{l}
  \cclrd{t}
  \;\cpar\;
  (  (\cclrd{r}\;\cpar\;(\cclrd{s}\,\cseq\,\cclrd{r}))
     \,\cseq\, 
     \cclrd{t} ) = 
\\
  \cclrd{t}
  \;\cpar\;
  (\cclrd{r}\,\cseq\, \cclrd{t})
  \;\cpar\;
  (\cclrd{s}\,\cseq\,\cclrd{r}\,\cseq\, \cclrd{t}) = 
\\
  \cclrd{s}\,\cseq\,\cclrd{r}\,\cseq\, \cclrd{t} 
\end{array}
\end{equation*}

\begin{equation*}
\begin{array}{l}
  (\cclrd{t}\;\cpar\;(\cclrd{r}\,\cseq\,\cclrd{t})) 
  \;\cpar\;
  (  \cclrd{s}
     \,\cseq\, 
     (\cclrd{t}\;\cpar\;(\cclrd{r}\,\cseq\,\cclrd{t})) ) = 
\\     
  \cclrd{t}\;\cpar\;(\cclrd{r}\,\cseq\,\cclrd{t})
  \;\cpar\;
  (\cclrd{s} \,\cseq\, \cclrd{t})
  \;\cpar\;
  (\cclrd{s} \,\cseq\, \cclrd{r}\,\cseq\,\cclrd{t})  = 
\\     
  \cclrd{s} \,\cseq\, \cclrd{r}\,\cseq\,\cclrd{t}
\end{array}
\end{equation*}

require

\begin{equation*}
\cclrd{r}\;\cpar\;(\cclrd{r} \,\cseq\, \cclrd{s}) = \cclrd{r} \,\cseq\, \cclrd{s}
\end{equation*}

>>>>

This appendix gives full typing derivations for the examplespresented in Chapter~\ref{ch:applications},
which demonstrate simple coeffect calculi for implicit parameters, liveness and data-flow.

% --------------------------------------------------------------------------------------------------

\section{Coeffect typing for implicit parameters}
\label{sec:appendixa-implicit}

The example on page~\pageref{pg:applications-flat-paramsex} considers the typing of a function that
adds the values of two implicit parameters: $\lambda x.\ident{?a} + \ident{?b}$. Note that addition
is a function (and so we write $(+)~\ident{?a}~\ident{?b}$) that requires no implicit parameters.
Thus our typing derivation starts with: 
%
\begin{equation*}
\Gamma_0 = (+):\ident{int}\xrightarrow{\cclrd{\emptyset}}\ident{int}\xrightarrow{\cclrd{\emptyset}}\ident{int}
\end{equation*}
%
The typing for the sub-expression $(+)~\ident{?a}$ is shown in Figure~\ref{fig:appendixa-impl} (a).
Note that the resulting function does not require any implicit parameters, so there is no non-determinism so far.
It is worth noting that this would change if we used $\eta$-expansion and wrote $(\lambda y.(+)~\ident{?a})$.
We refer to the above as $\Delta$. The rest of the expresion is shown in Figure~\ref{fig:appendixa-impl} (b).

Here, the last rule applies (\emph{app}) and so we non-deterministically split the set of required resources.
This means that we need $\cclrd{r_1}, \cclrd{r_2}$ such that $\cclrd{r_1}\,\cup\,\cclrd{r_2}=\{\ident{?a}:\ident{int}, \ident{?b}:\ident{int} \}$.
The Figure~\ref{fig:appendixa-impl} (c) summarizes the 9 options.

% --------------------------------------------------------------------------------------------------

\section{Coeffect typing for liveness}
\label{sec:appendixa-liveness}

\newcommand{\vrl}{\cclrd{\ident{L}}}
\newcommand{\vrd}{\cclrd{\ident{D}}}

The Section~\ref{sec:applications-flat-live} discusses a coeffect system where the coeffect 
annotations capture whether a variable may be accessed (marked as live using $\vrl$) or 
whether it is definitely not used (marked as dead using $\vrd$). The following three examples
are considered:
%
\begin{equation*}
\begin{array}{lll}
(\lambda x . 42)~y &~\hspace{1em}~&(1)\\
\ident{twoTimes}~42          &&(2)\\
(\lambda x . x)~42 &&(3)\\
\end{array}
\end{equation*}
%
The typing derivation for the expression ($1$) is shown in Figure~\ref{fig:appendixa-liv} (a).
The most interesting aspect about the previous example is the use of the (\emph{app}) rule, which
marks the resulting context as dead, even though a variable is accessed in the second part of the
expression (this part never needs to be evaluated).

Assuming $\Gamma_0 = \ident{twoTimes} : \ident{int} \xrightarrow{\vrl} \ident{int}$,
the typing derivation for ($2$) is shown in Figure~\ref{fig:appendixa-liv} (b).
Here, the variable context is marked as live. This is not because the argument of \ident{twoTimes} 
is marked as live, but because the function itself is a variable that (always) needs to be obtained
from the variable context. Finally, the derivation for ($3$) is shown in Figure~\ref{fig:appendixa-liv} (c).

% --------------------------------------------------------------------------------------------------

\section{Coeffect typing for data-flow}
\label{sec:appendixa-dataflow}

The Section~\ref{sec:applications-flat-dataflow} presents a coeffect type system that tracks the 
maximal number of past values required by a data-flow computation. The discussion includes the 
following example:
%
\begin{equation*}
\begin{array}{l}
(\,   \kvd{if}~~(\kvd{prev}~\ident{tick})=0\\[-0.25em]
\,\,\, \kvd{then}~(\lambda x \rightarrow \kvd{prev}~x)\\[-0.25em]
\,\,\, \kvd{else}~(\lambda x \rightarrow x) \,)\qquad(\kvd{prev}~\ident{counter})
\end{array}
\end{equation*}
%
In order to give typing for the above example, we first need to extend the language with 
conditionals. The typing rule for the $\kvd{if}$ construct is:
%
\begin{equation*}
\tyrule{if}
  { \coctx{\Gamma}{\cclrd{m}} \vdash e : \ident{bool} &
    \coctx{\Gamma}{\cclrd{n}} \vdash e_1 : \tau &
    \coctx{\Gamma}{\cclrd{n}} \vdash e_2 : \tau }
  {\coctx{\Gamma}{\textnormal{max}(\cclrd{m},\cclrd{n})} \vdash \kvd{if}~e~\kvd{then}~e_1~\kvd{else}~e_2 : \tau_1 }
\end{equation*}
%
Given a condition that requires $\cclrd{m}$ past and two (alternative) bodies that each require
$\cclrd{n}$ past values, the composed expression requires at most the maximum of the two, \ie~the
same context is passed to both the condition and the body using point-wise composition. As we will
see, the fact that both branches require the same number of past values means that we need to use
the sub-coeffecting rule.

Assuming $\Gamma_0 = \ident{tick}:\ident{int}, \ident{counter}:\ident{int}$, the 
Figure~\ref{fig:appendixa-dfl} shows the typing derivation for the example. In the first part,
we derive the types for the sub-expressions of the application and the conditional. Note that 
to obtain compatible function types, we use sub-coeffecting \emph{before} abstraction in the
typing of $(\lambda x.x)$ -- the type $\ident{int} \xrightarrow{\cclrd{1}} \ident{int}$ is  a
valid overapproximation. Finally, the typing of the application yields the requirement of two 
past values, calculated as $\textnormal{max}(\cclrd{1}, \cclrd{1}+\cclrd{1})$.

\newpage
% --------------------------------------------------------------------------------------------------

\begin{figure}
{\small a.) Typing for the sub-expression $(+)~\ident{?a}$}

\begin{equation*}
\begin{array}{l}
 \tyrule{app}
  { \hspace{-15em}\tyrule{var}{}
    { \coctx{\Gamma_0, x:\ident{int}}{\cclrd{ \emptyset }} \vdash
      (+) : \ident{int} \xrightarrow{\cclrd{\emptyset}} \ident{int} \xrightarrow{\cclrd{\emptyset}} \ident{int}} \\ \hspace{10em}
    \tyruler{param}{}
    { \coctx{\Gamma_0, x:\ident{int}}{\cclrd{ \{ \ident{?a}:\ident{int} \} }} \vdash
      \ident{?a} : \ident{int} } }
  { \coctx{\Gamma_0, x:\ident{int}}{\cclrd{ \{ \ident{?a}:\ident{int} \} }} \vdash
    (+)~\ident{?a} : \ident{int} \xrightarrow{\cclrd{\emptyset}} \ident{int} }
\end{array}
\end{equation*}

\vspace{1em}
{\small b.) Typing for the expression $(+)~\ident{?a}~\ident{?b}$}

\begin{equation*}
\tyrule{app}
{ \Delta\qquad\qquad &
  \coctx{\Gamma_0, x:\ident{int}}{\cclrd{ \{ \ident{?b}:\ident{int} \} }} \vdash
  \ident{?b} : \ident{int} }
{ \tyrule{abs}
  { \coctx{\Gamma_0, x:\ident{int}}{\cclrd{ \{\ident{?a}:\ident{int}, \ident{?b}:\ident{int} \} }} \vdash 
    (+)~\ident{?a}~\ident{?b} : \ident{int} }
  { \coctx{\Gamma_0}{\cclrd{ r_1 }} \vdash 
    \lambda x.(+)~\ident{?a}~\ident{?b} : 
    \ident{int} \xrightarrow{ \cclrd{r_2} } \ident{int} } }
\end{equation*}

\vspace{1em}
{\small c.) Possible splittings of the implicit parameters }

\begin{equation*}
\begin{array}{ll}
 \cclrd{r_1} = \{ \} & \cclrd{r_2} = \{ \ident{?a}:\ident{int}, \ident{?b}:\ident{int} \} \\
 \cclrd{r_1} = \{ \ident{?a}:\ident{int} \} & \cclrd{r_2} = \{ \ident{?a}:\ident{int}, \ident{?b}:\ident{int} \} \\
 \cclrd{r_1} = \{ \ident{?b}:\ident{int} \} & \cclrd{r_2} = \{ \ident{?a}:\ident{int}, \ident{?b}:\ident{int} \} \\
 \cclrd{r_1} = \{ \ident{?a}:\ident{int}, \ident{?b}:\ident{int} \} & \cclrd{r_2} = \{ \ident{?a}:\ident{int}, \ident{?b}:\ident{int} \} \\
 \cclrd{r_1} = \{ \ident{?a}:\ident{int}, \ident{?b}:\ident{int} \} & \cclrd{r_2} = \{ \}\\
 \cclrd{r_1} = \{ \ident{?a}:\ident{int}, \ident{?b}:\ident{int} \} & \cclrd{r_2} = \{ \ident{?a}:\ident{int} \}\\
 \cclrd{r_1} = \{ \ident{?a}:\ident{int}, \ident{?b}:\ident{int} \} & \cclrd{r_2} = \{ \ident{?b}:\ident{int} \}\\
 \cclrd{r_1} = \{ \ident{?a}:\ident{int} \} & \cclrd{r_2} = \{ \ident{?b}:\ident{int} \} \\
 \cclrd{r_1} = \{ \ident{?b}:\ident{int} \} & \cclrd{r_2} = \{ \ident{?a}:\ident{int} \} \\
\end{array}
\end{equation*}

\figcaption{Coeffect typing for implicit parameters}
\label{fig:appendixa-impl}
\end{figure}

% --------------------------------------------------------------------------------------------------

\begin{figure}[t]
{\small a.) Typing for the expression $(\lambda x . 42)~y$}

\begin{equation*}
\tyrule{app}
  { \hspace{-5em}
    \tyrule{abs}
      { \hspace{-5em}
        \tyrule{const}{}{ \coctx{y\!:\!\tau, x\!:\!\tau}{\vrd} \vdash 42 : \ident{int} } }
      { \coctx{y\!:\!\tau}{\vrd} \vdash \lambda x . 42 : \tau \xrightarrow{\vrd} \ident{int} }  &
    \tyrule{var}{}{\coctx{y\!:\!\tau}{\vrl} \vdash y : \tau } }
  { \inference 
    { \coctx{y\!:\!\tau}{\vrd \sqcup (\vrl \sqcap \vrd)} \vdash (\lambda x . 42)~y : \ident{int} }
    { \coctx{y\!:\!\tau}{\vrd} \vdash (\lambda x . 42)~y : \ident{int} } }
\end{equation*}

\vspace{1em}
{\small b.) Typing for the expression $\ident{twoTimes}~42$}

\begin{equation*}
\tyrule{app}
  { \hspace{-5em}
    \tyrule{var}{}{ \coctx{\Gamma_0}{\vrl} \vdash \ident{twoTimes} : \ident{int} \xrightarrow{\vrl} \ident{int}  }&
    \tyrule{var}{}{ \coctx{\Gamma_0}{\vrd} \vdash 42 : \ident{int} } }
  { \inference 
      { \coctx{\Gamma_0}{\vrl \sqcup (\vrd \sqcap \vrl)} \vdash \ident{twoTimes}~42 : \ident{int}}
      { \coctx{\Gamma_0}{\vrl} \vdash \ident{twoTimes}~42 : \ident{int}} }
\end{equation*}

\vspace{1em}
{\small c.) Typing for the expression $(\lambda x . x)~42$}

\begin{equation*}
\tyrule{app}
  { \hspace{-5em}
    \tyrule{abs}
      { \hspace{-5em}
        \tyrule{var}{}{ \coctx{x\!:\!\ident{int}}{\vrl} \vdash x : \ident{int} } }
      { \coctx{()}{\vrl} \vdash \lambda x . x : \ident{int} \xrightarrow{\vrl} \ident{int} }  &
    \tyrule{const}{}{\coctx{()}{\vrd} \vdash 42 : \ident{int} } }
  { \inference 
    { \coctx{()}{\vrl \sqcup (\vrl \sqcap \vrd)} \vdash (\lambda x . x)~42 : \ident{int} }
    { \coctx{()}{\vrl} \vdash (\lambda x . x)~42 : \ident{int} } }
\end{equation*}

\figcaption{Coeffect typing for liveness}
\label{fig:appendixa-liv}
\end{figure}

% --------------------------------------------------------------------------------------------------

\begin{figure}[t]
{\small a.) Typing for sub-expressions of the conditional and for the argument}

\begin{equation*}
\tyrule{prev}
 { \coctx{\Gamma_0}{\cclrd{0}} \vdash \ident{tick} : \ident{int} }
 { \inference
   { \coctx{\Gamma_0}{\cclrd{1}} \vdash \kvd{prev}~\ident{tick} : \ident{int} }
   { \coctx{\Gamma_0}{\cclrd{1}} \vdash (\kvd{prev}~\ident{tick}) = 0 : \ident{bool} } }
\end{equation*}
\vspace{1em}
\begin{equation*}
\tyrule{prev}
 { \tyrule{var}
    {} 
    { \coctx{\Gamma_0,x\!:\!\ident{int}}{\cclrd{0}} \vdash x : \ident{int} } }
 { \tyrule{abs}
   { \coctx{\Gamma_0,x\!:\!\ident{int}}{\cclrd{1}} \vdash \kvd{prev}~x : \ident{int} }
   { \coctx{\Gamma_0}{\cclrd{1}} \vdash \lambda x . \kvd{prev}~x : \ident{int} \xrightarrow{\cclrd{1}} \ident{int} } }
\end{equation*}
\vspace{1em}
\begin{equation*}
\tyrule{abs}
  { \tyrule{sub}
      { \tyrule{var}{}{ \coctx{\Gamma_0, x\!:\!\ident{int}}{\cclrd{0}} \vdash x : \ident{int} } }
      { \coctx{\Gamma_0, x\!:\!\ident{int}}{\cclrd{1}} \vdash x : \ident{int} } } 
  { \coctx{\Gamma_0}{\cclrd{0}} \vdash \lambda x . x : \ident{int} \xrightarrow{\cclrd{1}} \ident{int} }
\end{equation*}
\vspace{1em}
\begin{equation*}
\tyrule{prev}
  { \tyrule{var}{}{ \coctx{\Gamma_0}{\cclrd{0}} \vdash \ident{counter} : \ident{int} } }
  { \coctx{\Gamma_0}{\cclrd{1}} \vdash \kvd{prev}~\ident{counter} : \ident{int} }
\end{equation*}

\vspace{1em}
{\small b.) Typing for the composed expression}

\begin{equation*}
\tyrule{app}
  { \hspace{-5em}
    \tyrule{if}
      { (\ldots) } 
      { \coctx{\Gamma_0}{\cclrd{1}} \vdash (\kvd{if}\ldots\kvd{then}\ldots\kvd{else}\ldots) : \ident{int} \xrightarrow{\cclrd{1}} \ident{int} } & 
        \coctx{\Gamma_0}{\cclrd{1}} \vdash (\ldots) : \ident{int} }
  { \hspace{-1em}
    \inference
     { \coctx{\Gamma_0}{\textnormal{max}(\cclrd{1}, \cclrd{1}+\cclrd{1})} \vdash (\kvd{if}\ldots\kvd{then}\ldots\kvd{else}\ldots)~(\ldots) : \ident{int} }
     { \coctx{\Gamma_0}{\cclrd{2}} \vdash (\kvd{if}\ldots\kvd{then}\ldots\kvd{else}\ldots)~(\ldots) : \ident{int} } }
\end{equation*}

\figcaption{Coeffect typing for data-flow}
\label{fig:appendixa-dfl}
\end{figure}

% --------------------------------------------------------------------------------------------------

### 2014-12-07 text/Chapter01.tex
Finish tweaks in flat coeffects (Chapter 3)

\caption{Conditional compilation in the HTTP module of the F\# Data library}
\label{fig:introduction-context-http}
\end{figure}

@@

\caption{Client/server news reader app implemented in a context-aware language}
\label{fig:introduction-sample}
\end{figure}

>>>>

\figcaption{Conditional compilation in the HTTP module of the F\# Data library}
\label{fig:introduction-context-http}
\end{figure}

@@

\figcaption{Client/server news reader app implemented in a context-aware language}
\label{fig:introduction-sample}
\end{figure}

### 2014-12-07 text/Chapter02.tex
Finish tweaks in flat coeffects (Chapter 3)

\caption{Simple effect system}
\label{fig:path-eff}
\end{figure}

@@

\caption{Simple coeffect system}
\label{fig:path-coeff}
\end{figure}

@@

\caption{Typing for a comonadic language with contextual staged computations}
\label{fig:modal-meta}
\end{figure}

@@

\caption{Exchange, weakening and contraction typing rules}
\label{fig:substructural-rules}
\end{figure}

@@

\begin{figure}
\begin{equation*}
\tyrule{exchange1}
  {\Gamma(\Delta, \Sigma) \vdash e : \alpha}
  {\Gamma(\Sigma, \Delta) \vdash e : \alpha}
\end{equation*}
\begin{equation*}
\tyrule{exchange2}
  {\Gamma(\Delta; \Sigma) \vdash e : \alpha}
  {\Gamma(\Sigma; \Delta) \vdash e : \alpha}
\end{equation*}
\begin{equation*}
\tyrule{weakening}
  {\Gamma(\Delta) \vdash e : \alpha}
  {\Gamma(\Delta; \Sigma) \vdash e : \alpha}
\end{equation*}
\begin{equation*}
\tyrule{contraction}
  {\Gamma(\Delta; \Sigma) \vdash e : \alpha}
  {\Gamma(\Delta) \vdash \subst{e}{\Sigma}{\Delta} : \alpha}
\end{equation*}
\caption{Exchange, weakening and contraction rules for bunched typing}
\label{fig:substructural-bunched}
\end{figure}

>>>>

\figcaption{Simple effect system}
\label{fig:path-eff}
\end{figure}

@@

\figcaption{Simple coeffect system}
\label{fig:path-coeff}
\end{figure}

@@

\figcaption{Typing for a comonadic language with contextual staged computations}
\label{fig:modal-meta}
\end{figure}

@@

\figcaption{Exchange, weakening and contraction typing rules}
\label{fig:substructural-rules}
\end{figure}

@@

\begin{figure}
\begin{equation*}
\tyrule{exchange1}
  {\Gamma(\Delta, \Sigma) \vdash e : \alpha}
  {\Gamma(\Sigma, \Delta) \vdash e : \alpha}
\end{equation*}
\begin{equation*}
\tyrule{exchange2}
  {\Gamma(\Delta; \Sigma) \vdash e : \alpha}
  {\Gamma(\Sigma; \Delta) \vdash e : \alpha}
\end{equation*}
\begin{equation*}
\tyrule{weakening}
  {\Gamma(\Delta) \vdash e : \alpha}
  {\Gamma(\Delta; \Sigma) \vdash e : \alpha}
\end{equation*}
\begin{equation*}
\tyrule{contraction}
  {\Gamma(\Delta; \Sigma) \vdash e : \alpha}
  {\Gamma(\Delta) \vdash \subst{e}{\Sigma}{\Delta} : \alpha}
\end{equation*}
\figcaption{Exchange, weakening and contraction rules for bunched typing}
\label{fig:substructural-bunched}
\end{figure}
